Dynamic Epistemic Logic - Final Take Home Exam

### 2021-07-03 del/final_exam.tex
DEL: Final Exam: Exercises 4 to 6, version that was send after the deadline

\usepackage{amsmath,latexsym,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures_final_exam/} }
% Required packages for plotting
\usepackage{pgfplots}
%\usepackage{pgfplotstable}
\pgfplotsset{compat=1.16}
%\usepackage{tikz}

@@

We will show that $M,s \models \left<\phi\right> K_a \psi$ iff $M,s \models (\phi \wedge K_a(\phi \rightarrow \left<\phi\right>\psi))$
\begin{flalign*}
	& M,s \models (\phi \wedge K_a(\phi \rightarrow \left<\phi\right>\psi))\\
	\Leftrightarrow & M,s \models \phi \text{ and } M,s \models K_a(\phi \rightarrow \left<\phi\right>\psi)\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \rightarrow \left<\phi\right>\psi]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \text{ implies } M,t \models \left<\phi\right>\psi]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \text{ implies } (M,t \models \phi \text{ and } M|_\phi,t \models \psi)]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \text{ implies } M|_\phi,t \models \psi]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]: M,t \models \phi) \text{ implies } M|_\phi,t \models \psi\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]|_\phi) \text{ implies } M|_\phi,t \models \psi\\
	\Leftrightarrow & M,s \models \phi \text{ and } M|_\phi,s \models K_a\psi\\
	\Leftrightarrow & M,s \models \left<\phi\right> K_a\psi
\end{flalign*}

\subsection*{Exercise 4.21}
In Example 4.1 from the book we can first see that the formula $\phi = \left<p \wedge \neg K_b p\right>\hat{K}_a\hat{K}_b \neg p$ is not valid for the $0$ world since
\[M,0 \not\models p \Rightarrow M,0 \not\models p \wedge \neg K_b p\]
Then for the $1$ world we have that $M,1 \models p$, but also $M,1 \models \neg K_b p$ since $\neg K_b p \leftrightarrow \hat{K}_b \neg p$ and indeed $b$ considers possible the $0$ world where $\neg p$ holds.
Then we see that $M,1 \models \hat{K}_a\hat{K}_b \neg p$, because
\begin{align*}
	& M,0 \models \neg p \\
	\Rightarrow & M,1 \models \hat{K}_b \neg p & \text{ (since } 0 \sim_b 1 \text{)}\\
	\Rightarrow & M,1 \models \hat{K}_a \hat{K}_b \neg p & \text{ (since } 1 \sim_a 1 \text{)}
\end{align*}
Therefore finally we have
\[M,1 \models p \text{ and } M,1 \models \neg K_b p \text{ and } M,1 \models \hat{K}_a\hat{K}_b \neg p \Rightarrow M,1 \models \left<p \wedge \neg K_b p\right>\hat{K}_a\hat{K}_b \neg p\]

>>>>

\usepackage{amsmath,latexsym,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures_final_exam/} }
% Required packages for plotting
\usepackage{pgfplots}
%\usepackage{pgfplotstable}
\pgfplotsset{compat=1.16}
%\usepackage{tikz}
\usepackage{multicol}
\usepackage{logicproof}

@@

We will show that $M,s \models \left<\phi\right> K_a \psi$ iff $M,s \models (\phi \land K_a(\phi \rightarrow \left<\phi\right>\psi))$
\begin{flalign*}
	& M,s \models (\phi \land K_a(\phi \rightarrow \left<\phi\right>\psi))\\
	\Leftrightarrow & M,s \models \phi \text{ and } M,s \models K_a(\phi \rightarrow \left<\phi\right>\psi)\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \rightarrow \left<\phi\right>\psi]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \text{ implies } M,t \models \left<\phi\right>\psi]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \text{ implies } (M,t \models \phi \text{ and } M|_\phi,t \models \psi)]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]) [M,t \models \phi \text{ implies } M|_\phi,t \models \psi]\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]: M,t \models \phi) \text{ implies } M|_\phi,t \models \psi\\
	\Leftrightarrow & M,s \models \phi \text{ and } (\forall t \in R[s]|_\phi) \text{ implies } M|_\phi,t \models \psi\\
	\Leftrightarrow & M,s \models \phi \text{ and } M|_\phi,s \models K_a\psi\\
	\Leftrightarrow & M,s \models \left<\phi\right> K_a\psi
\end{flalign*}

\subsection*{Exercise 4.21}
In Example 4.1 from the book we can first see that the formula $\phi = \left<p \land \neg K_b p\right>\hat{K}_a\hat{K}_b \neg p$ is not valid for the $0$ world since
\[M,0 \not\models p \Rightarrow M,0 \not\models p \land \neg K_b p\]
Then for the $1$ world we have that $M,1 \models p$, but also $M,1 \models \neg K_b p$ since $\neg K_b p \leftrightarrow \hat{K}_b \neg p$ and indeed $b$ considers possible the $0$ world where $\neg p$ holds.
Then we see that $M,1 \models \hat{K}_a\hat{K}_b \neg p$, because
\begin{align*}
	& M,0 \models \neg p \\
	\Rightarrow & M,1 \models \hat{K}_b \neg p & \text{ (since } 0 \sim_b 1 \text{)}\\
	\Rightarrow & M,1 \models \hat{K}_a \hat{K}_b \neg p & \text{ (since } 1 \sim_a 1 \text{)}
\end{align*}
Therefore finally we have
\[M,1 \models p \text{ and } M,1 \models \neg K_b p \text{ and } M,1 \models \hat{K}_a\hat{K}_b \neg p \Rightarrow M,1 \models \left<p \land \neg K_b p\right>\hat{K}_a\hat{K}_b \neg p\]

\section*{Exercise 4}

We will use induction on the formula $\phi$.

\paragraph{if $\phi = p$} then $\phi(p/\psi) = \psi$ and $\phi(p/\chi) = \chi$ and finally $\vdash \psi \leftrightarrow \chi$ from hypothesis

\paragraph{if $\phi = \neg \phi^\prime$} then $\phi(p/\psi) = \neg \phi^\prime(p/\psi)$ and $\phi(p/\chi) = \neg \phi^\prime(p/\chi)$.
So we have
\begin{logicproof}{1}
	\vdash \phi^\prime(p/\psi) \leftrightarrow \phi^\prime(p/\chi) & (induction hypothesis) \\
	\vdash \neg \phi^\prime(p/\psi) \leftrightarrow \neg \phi^\prime(p/\chi) & (1, propositional) \\
	\vdash \phi(p/\psi) \leftrightarrow \phi(p/\chi) &
\end{logicproof}

\paragraph{if $\phi = \phi_1 \land \phi_2$} then $\phi(p/\psi) = \phi_1(p/\psi) \land \phi_2(p/\psi)$ and $\phi(p/\chi) = \phi_1(p/\chi) \land \phi_2(p/\chi)$.\\
So we have
\begin{logicproof}{1}
	\vdash \phi_1(p/\psi) \leftrightarrow \phi_1(p/\chi) & (induction hypothesis) \\
	\vdash \phi_2(p/\psi) \leftrightarrow \phi_2(p/\chi) & (induction hypothesis) \\
	\vdash \phi_1(p/\psi) \land \phi_2(p/\psi) \leftrightarrow \phi_1(p/\chi) \land \phi_2(p/\chi) & (1,2, propositional) \\
	\vdash \phi(p/\psi) \leftrightarrow \phi(p/\chi) &
\end{logicproof}

\paragraph{if $\phi = K_a \phi^\prime$} then $\phi(p/\psi) = K_a \phi^\prime(p/\psi)$ and $\phi(p/\chi) = K_a \phi^\prime(p/\chi)$.\\
So we have
\begin{logicproof}{1}
	\vdash \phi^\prime(p/\psi) \leftrightarrow \phi^\prime(p/\chi) & (induction hypothesis) \\
	\vdash K_a(\phi^\prime(p/\psi) \leftrightarrow \phi^\prime(p/\chi)) & (1, necessitation of $K_a$) \\
	\vdash K_a(\phi^\prime(p/\psi) \rightarrow \phi^\prime(p/\chi)) & (2, propositional) \\
	\vdash K_a(\phi^\prime(p/\psi)) \rightarrow K_a(\phi^\prime(p/\chi)) & (3, distribution of $K_a$ over $\rightarrow$) \\
	\vdash K_a(\phi^\prime(p/\chi) \rightarrow \phi^\prime(p/\psi)) & (2, propositional) \\
	\vdash K_a(\phi^\prime(p/\chi)) \rightarrow K_a(\phi^\prime(p/\psi)) & (5, distribution of $K_a$ over $\rightarrow$) \\
	\vdash K_a(\phi^\prime(p/\psi)) \leftrightarrow K_a(\phi^\prime(p/\chi)) & (4,6, propositional) \\
	\vdash \phi(p/\psi) \leftrightarrow \phi(p/\chi) &
\end{logicproof}

\paragraph{if $\phi = [\phi_1]\phi_2$} then $\phi(p/\psi) = [\phi_1(p/\psi)]\phi_2(p/\psi)$ and $\phi(p/\chi) = [\phi_1(p/\chi)]\phi_2(p/\chi)$.
\\[8pt]
Here we will also use the following tautology (for arbitrary $\phi_1,\phi_2,\psi,\chi$)
\begin{equation}
	((\phi_1 \rightarrow \psi) \land (\phi_2 \rightarrow \chi) \land (\phi_1 \leftrightarrow \phi_2)) \rightarrow (\psi \leftrightarrow \chi) \label{t1}
\end{equation}
So we have
\begin{logicproof}{1}
	\vdash \phi_2(p/\psi) \rightarrow [\phi_1(p/\psi)]\phi_2(p/\psi) & (from $\phi$ follows $[\psi]\phi$) \\
	\vdash \phi_2(p/\chi) \rightarrow [\phi_1(p/\chi)]\phi_2(p/\chi) & (from $\phi$ follows $[\psi]\phi$) \\
	\vdash \phi_2(p/\psi) \leftrightarrow \phi_2(p/\chi) & (inductive hypothesis) \\
	\vdash [\phi_1(p/\psi)]\phi_2(p/\psi) \leftrightarrow [\phi_1(p/\chi)]\phi_2(p/\chi) & (1,2,3, tautology \eqref{t1}, modus ponence) \\
	\vdash \phi(p/\psi) \leftrightarrow \phi(p/\chi) &
\end{logicproof}
\newpage

\section*{Exercise 5}

We have the following models:
\begin{multicols}{2}
	\[M = (\text{Read};\text{Read}_b)=\left<S,\sim,\textbf{pre}\right>\]
	\columnbreak
	\begin{center}
		\begin{tikzpicture}[
				node distance={20mm},
				shorten >=1pt,
				shorten <=1pt,
				main/.style = {draw, circle, inner sep=5pt}
			]
			\node[main, label={(np,npb)}, label=left:{$\neg p$}] (nn) {};
			\node[main, label={(np,pb)}, label=right:{$\bot$}] (np) [right of=nn] {};
			\node[main, label=below:{(p,npb)}, label=left:{$\bot$}] (pn) [below of=nn] {};
			\node[main, label=below:{(p,pb)}, label=right:{$p$}] (pp) [right of=pn] {};
			\draw[-] (nn) to node[above] {$a$} (np);
			\draw[-] (nn) to node[left] {$b$} (pn);
			\draw[-] (np) to node[right] {$b$} (pp);
			\draw[-] (pn) to node[below] {$a$} (pp);
		\end{tikzpicture}
	\end{center}
\end{multicols}
\begin{multicols}{2}
	\[M^\prime = (\text{Read};\text{Read}_b)^\prime=\left<S^\prime,\sim^\prime,\textbf{pre}^\prime\right>\]
	\columnbreak
	\begin{center}
		\begin{tikzpicture}[
				node distance={20mm},
				shorten >=1pt,
				shorten <=1pt,
				main/.style = {draw, circle, inner sep=5pt}
			]
			\node[main, label={(np,npb)}, label=below:{$\neg p$}] (nn) {};
			\node[main, label={(p,pb)}, label=below:{$p$}] (pp) [right of=nn] {};
		\end{tikzpicture}
	\end{center}
\end{multicols}
In order to show that $M$ is emulus to $M^\prime$ we will use the definition in footnote 4 from p.159 of the book.
We consider the relation
\[\mathfrak{E} = S \times S^\prime\]
We will show that $\mathfrak{E}$ is an emulation between $(M,(\text{np,npb}))$ and $(M^\prime,(\text{np,npb}))$.
We indeed have $\mathfrak{E}((\text{np,npb}),(\text{np,npb}))$

\paragraph{Front}
\begin{itemize}
	\item For $\mathfrak{E}((\text{np,npb}),(\text{np,npb}))$ and $(\text{np,npb}) \sim_a (\text{np,pb})$ we have\\
		$(\text{np,npb}) \sim^\prime_a (\text{np,npb})$ and $\mathfrak{E}((\text{np,npb}),(\text{np,pb}))$
	\item For $\mathfrak{E}((\text{np,npb}),(\text{np,npb}))$ and $(\text{np,npb}) \sim_b (\text{p,npb})$ we have\\
		$(\text{np,npb}) \sim^\prime_b (\text{np,npb})$ and $\mathfrak{E}((\text{np,npb}),(\text{p,npb}))$
	\item For $\mathfrak{E}((\text{np,pb}),(\text{np,npb}))$ and $(\text{np,pb}) \sim_b (\text{p,pb})$ we have\\
		$(\text{np,pb}) \sim^\prime_b (\text{np,npb})$ and $\mathfrak{E}((\text{np,npb}),(\text{p,pb}))$
	\item For $\mathfrak{E}((\text{p,npb}),(\text{np,npb}))$ and $(\text{p,npb}) \sim_a (\text{p,pb})$ we have\\
		$(\text{p,npb}) \sim^\prime_a (\text{np,npb})$ and $\mathfrak{E}((\text{np,npb}),(\text{p,pb}))$
\end{itemize}

\paragraph{Back}
For each $\text{t}^\prime \in S^\prime$ we only have $\text{t}^\prime \sim^\prime_a \text{t}^\prime$ and $\text{t}^\prime \sim^\prime_b \text{t}^\prime$ so for $\mathfrak{E}(\text{t},\text{t}^\prime)$ we selecting $\text{t}$ again suffices since also $\text{t} \sim_a \text{t}$ and $\text{t} \sim_b \text{t}$

\paragraph{Pre}
For all $\mathfrak{E}(\text{t},(\text{np,npb})), \text{t} \in S$ iy holds that there are $(\text{np,npb}),(\text{p,pb}) \in S^\prime$ where
\[\textbf{pre}(\text{t}) \models \textbf{pre}^\prime(\text{np,npb}) \lor \textbf{pre}^\prime(\text{p,pb}) \Rightarrow \textbf{pre}(\text{t}) \models p \lor \neg p \Rightarrow \textbf{pre}(\text{t}) \models \top\]
which holds for all $\text{t} \in S$. For the reverse we only need to select t and (np,npb) and
\[\textbf{pre}^\prime(\text{np,npb}) \models \text{pre(t)} \lor \textbf{pre}(\text{np,npb}) \Rightarrow \neg p \models \text{pre(t)} \lor \neg p\]
which holds, and therefore $\mathfrak{E}$ is an emulation between $(M,(\text{np,npb}))$ and $(M^\prime,(\text{np,npb}))$.
\\[8pt]
Likewise we can show that $\mathfrak{E}$ is an emulation between $(M,(\text{p,pb}))$ and $(M^\prime,(\text{p,pb}))$ and therefore $\mathfrak{E}$ is a total emulation.

\section*{Exercise 6}

We will use the Example 6.38 of the book where $\vdash_{\textbf{AMC}} [\text{Read, p}]K_a p$.
We also note that using the same logic we get $\vdash_{\textbf{AMC}} [\text{Read, np}]K_a \neg p$.\\
So we have
\begin{footnotesize}
\begin{logicproof}{1}
	\vdash_{\textbf{AMC}} [\text{Read, p}]K_a p & (Example 6.38) \\
	\vdash_{\textbf{AMC}} [\text{Read, p}](K_a p \lor K_a \neg p) & (1, propositional) \\
	\vdash_{\textbf{AMC}} K_b[\text{Read, p}](K_a p \lor K_a \neg p) & (2, necessitation of $K_b$) \\
	\vdash_{\textbf{AMC}} p \rightarrow K_b[\text{Read, p}](K_a p \lor K_a \neg p) & (3, weakening) \\
	\vdash_{\textbf{AMC}} [\text{Read, np}]K_a \neg p & (Example 6.38 for np) \\
	\vdash_{\textbf{AMC}} [\text{Read, np}](K_a p \lor K_a \neg p) & (5, propositional) \\
	\vdash_{\textbf{AMC}} K_b[\text{Read, p}](K_a p \lor K_a \neg p) & (6, necessitation of $K_b$) \\
	\vdash_{\textbf{AMC}} p \rightarrow K_b[\text{Read, p}](K_a p \lor K_a \neg p) & (7, weakening) \\
	\vdash_{\textbf{AMC}} [\text{Read, p}]K_b(K_a p \lor K_a \neg p) \leftrightarrow (p \rightarrow \bigwedge_{\text{s} \sim_b \text{t}} K_b[\text{Read, t}](K_a p \lor K_a \neg p)) & (pre(p) $=p$, action and knowledge) \\
	\vdash_{\textbf{AMC}} [\text{Read, p}]K_b(K_a p \lor K_a \neg p) & ($[\text{p}]_{\sim_b} = \{\text{p,np}\}$, 4,8,9, propositional)
\end{logicproof}
\end{footnotesize}
